\PassOptionsToPackage{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}{hyperref}
\documentclass[11pt]{article}

% Encoding, fonts, spacing
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{setspace}
\onehalfspacing

% Page layout (adjust margins as you prefer)
%\usepackage[margin=1in]{geometry}

% Math, symbols, tables, graphics
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{bookmark}
\usepackage{graphicx}

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

% URLs, hyperlinks, bibliography
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}

% Authors and affiliations
\usepackage{authblk}

% Microtypography
\usepackage{microtype}

% TikZ for diagrams
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{arrows.meta,calc,fit,positioning,shapes.geometric}

% Framed callouts
\usepackage{mdframed}
\usepackage[most]{tcolorbox}

% Simple keywords helper
\providecommand{\keywords}[1]{\textbf{Keywords: } #1}

% Canonicalization macro helpers
\newcommand{\Raw}{\mathsf{Raw}}
\newcommand{\Canon}{\mathsf{Canon}}

% Figure callout box
\newcommand{\FigureCallout}[2]{%
  \begin{tcolorbox}[
    colback=gray!5,
    colframe=black!40,
    title={#1},
    fonttitle=\bfseries,
    arc=3pt,
    boxrule=0.5pt,
    width=\linewidth,
    enhanced,
    breakable
  ]
  #2
  \end{tcolorbox}
}

% Paragraph formatting
\setlength{\parskip}{0.75em}
\setlength{\parindent}{0em}

% Itemized list formatting
\usepackage{enumitem}
\setlist[itemize]{itemsep=0.2em, topsep=0.2em, parsep=0em, partopsep=0em}

% Define missing commands
\providecommand{\Raw}{\mathrm{Raw}}
\providecommand{\Canon}{\mathrm{Canon}}

\title{Categorical Semantics for the Civic Exchange Protocol (CEP)}

\author[1,2]{Denise Case}
\affil[1]{Northwest Missouri State University, Computer Science and Information Systems, Maryville, MO, USA}
\affil[2]{Civic Interconnect, Ely, MN, USA}

\date{\today}

\begin{document}

\maketitle
\vspace{-1em}

\begin{tcolorbox}[colback=gray!10, colframe=black!20, boxrule=0.3pt]
    \textbf{Preprint Notice.}
    This preprint has not undergone peer review.
    It is shared to support community discussion, transparency research, and early technical evaluation.
\end{tcolorbox}

\begin{abstract}
The Civic Exchange Protocol (CEP) defines entities, relationships, exchanges,
and context tags as foundational primitives for civic information systems.
This paper develops a categorical semantics for CEP, providing a rigorous
mathematical basis for its compositional structure, identity guarantees,
and interoperability claims.
CEP records are modeled as typed objects in a finitely complete category,
with morphisms representing provenance-preserving transformations.
Envelopes, attestations, and context tags arise as natural transformations
between functors encoding system-level views of the data.
Canonicalization is formulated as a strict monoidal functor that preserves
identity and equivalence classes, while jurisdictional adapters appear as
oplax functors mediating between local schemas and global vocabularies.
Together, these constructions clarify CEP's invariants, formalize correctness
properties of its identifier mechanism (SNFEI),
and establish a robust foundation for future work in verification,
data fusion, and cross-jurisdiction interoperability.

\end{abstract}

\begin{keywords}
    Civic data;
    category theory;
    functorial data modeling;
    interoperability;
    canonicalization;
    identifiers;
    provenance.
\end{keywords}

\input{01_intro}
\input{02_preliminaries}
\input{03_cep}
\input{04_nt}
\input{05_canonicalization}
\input{06_adapters}
\input{07_ctags}
\input{08_results}
\input{09_apps}
\input{70_limitations}
\input{90_conclusion}

\appendix
\clearpage

\input{99_appendixA}   % Category background
\input{99_appendixB}   % Worked examples
\input{99_appendixC}   % Proof sketches
\input{99_appendixD}   % Diagrammatic intuition
\input{99_appendixE}   % Glossary of categorical notions

\bibliographystyle{plainnat}
\bibliography{00_cep}  %  .bib file name

\end{document}
