% ============================================================
\section{Interoperability Results}
\label{sec:results}
% ============================================================

The categorical semantics developed in the preceding sections yields
formal guarantees for CEP's central interoperability claims. These
results follow from functoriality, naturality, monoidal structure, and
the oplax semantics of adapters.

% ------------------------------------------------------------
\subsection{Functorial Consistency}
% ------------------------------------------------------------

We first show that the interaction between record evolution and
attestation is coherent under composition.

\begin{lemma}[Functorial Consistency of Attested Evolutions]
\label{lemma:functorial-consistency}
Let $f : R \to R'$ and $g : R' \to R''$ be morphisms in
$\mathbf{CEP}$, and let $\alpha : \mathcal{E} \Rightarrow \mathcal{E}'$
be the natural transformation representing attestation.
Then the following diagram commutes:
\[
\mathcal{E}'(g \circ f)
  \;=\;
\alpha_{R''} \circ \mathcal{E}(g \circ f)
  \;=\;
\mathcal{E}'(g) \circ \mathcal{E}'(f).
\]
\end{lemma}

\begin{proof}
By naturality of $\alpha$, each attestation map satisfies
$\alpha_{R'} \circ \mathcal{E}(f) = \mathcal{E}'(f) \circ \alpha_{R}$.
Associativity of morphism composition in $\mathbf{CEP}$ gives the result.
\end{proof}

This establishes that CEP's update–attestation workflow is
compositionally well-behaved, enabling reliable audit chains and
provable immutability of provenance.

% ------------------------------------------------------------
\subsection{Identifier Preservation}
% ------------------------------------------------------------

We now formalize the guarantee that canonical identifiers remain stable
under all valid record evolutions.

\begin{theorem}[Identifier Preservation]
\label{thm:identifier-preservation}
Let $f : R \to R'$ be any morphism in $\mathbf{CEP}$.
Let $\mathcal{C}$ denote the canonicalization functor and $H$ the
deterministic hash used to compute the SNFEI.
Then:
\[
H(\mathcal{C}(R)) \;=\; H(\mathcal{C}(R')).
\]
\end{theorem}

\begin{proof}
By definition of a valid $\mathbf{CEP}$ morphism, $f$ preserves the
canonical components of the record. By the strict monoidality of
$\mathcal{C}$, we have $\mathcal{C}(R) = \mathcal{C}(R')$. Since $H$ is
a pure function, the SNFEI values coincide.
\end{proof}

This result provides the mathematical foundation for CEP's claim of
\emph{identity stability across revisions}, ensuring that record
evolution does not alter canonical identity.

% ------------------------------------------------------------
\subsection{Cross-Jurisdiction Reconciliation}
% ------------------------------------------------------------

Jurisdictional adapters mediate between heterogeneous local structures
and the global CEP vocabulary. We show that their oplax semantics
preserves canonical equivalence classes.

\begin{theorem}[Canonical Equivalence Preservation Under Adapters]
\label{thm:adapter-preservation}
Let $\mathcal{A}_1, \mathcal{A}_2$ be valid jurisdictional adapters
(as defined in Section~\ref{sec:adapters}).  
If two local records $R$ and $R'$ satisfy
\[
H(\mathcal{C}(\mathcal{A}_1(R)))
  \;=\;
H(\mathcal{C}(\mathcal{A}_1(R'))),
\]
then after composition with a second adapter,
\[
H(\mathcal{C}(\mathcal{A}_2(\mathcal{A}_1(R))))
  \;=\;
H(\mathcal{C}(\mathcal{A}_2(\mathcal{A}_1(R')))).
\]
\end{theorem}

\begin{proof}
Each adapter $\mathcal{A}_i$ is oplax, meaning it may weaken local
structure but cannot introduce new canonical content (Section~6.3).
Thus $\mathcal{C}(\mathcal{A}_i(R))$ is defined and stable. The strict
monoidality of $\mathcal{C}$ ensures that canonical strings are preserved
under composition of such structure-weakening functors. Applying $H$
yields the claimed equality.
\end{proof}

\noindent
This theorem establishes that cross-jurisdiction data pipelines cannot
violate CEP's canonical identity invariants, provided the adapters
conform to the correctness criteria of Section~\ref{sec:adapters}.  
It provides the formal backbone for CEP's claim of
\emph{globally consistent, locally autonomous interoperability}.

