\PassOptionsToPackage{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}{hyperref}
% REQ-FILE: The above line, \PassOptionsToPackage{} must be very first line.

\documentclass[11pt]{article}

% Encoding and font setup for cross-platform reproducibility
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Improve readability for dense theoretical material
\usepackage{setspace}
\onehalfspacing

% Mathematical notation and table formatting for formal semantics
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{bookmark}
\usepackage{graphicx}

% Theorem environments for definitions, lemmas, and formal statements
\usepackage{amsthm}

% plain style is italic body text (for theorems, lemmas, etc.)
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

% definition style = upright body text (for definitions, examples)
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

% remark style = upright, lighter weight (for remarks, notes)
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{note}[theorem]{Note}

% Bibliography management and hyperlink support
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}

% Support multiple author affiliations
\usepackage{authblk}

% Improve typographic quality and line breaking
\usepackage{microtype}

% Categorical and commutative diagrams
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{arrows.meta,calc,fit,positioning,shapes.geometric}

% Callout boxes for examples and emphasis
\usepackage{mdframed}
\usepackage[most]{tcolorbox}

% Keywords macro for structured abstract metadata
\providecommand{\keywords}[1]{\textbf{Keywords: } #1}

% Notation for Raw vs Canonical categories in CEP/CEE
\newcommand{\Raw}{\mathsf{Raw}}
\newcommand{\Canon}{\mathsf{Canon}}

% Reusable figure callout box for visual emphasis
\newcommand{\FigureCallout}[2]{%
  \begin{tcolorbox}[
    colback=gray!5,
    colframe=black!40,
    title={#1},
    fonttitle=\bfseries,
    arc=3pt,
    boxrule=0.5pt,
    width=\linewidth,
    enhanced,
    breakable
  ]
  #2
  \end{tcolorbox}
}

% Paragraph spacing prioritizes readability over traditional indentation
\setlength{\parskip}{0.75em}
\setlength{\parindent}{0em}

% Compact list formatting for dense conceptual content
\usepackage{enumitem}
\setlist[itemize]{itemsep=0.2em, topsep=0.2em, parsep=0em, partopsep=0em}

% Defensive re-definition ensures commands exist if loaded earlier
\providecommand{\Raw}{\mathrm{Raw}}
\providecommand{\Canon}{\mathrm{Canon}}

\title{A Formal Ontology for Civic Accountable Entities (CAE)}

\author[1,2]{Denise M. Case}
\affil[1]{Northwest Missouri State University, Computer Science and Information Systems, Maryville, MO, USA}
\affil[2]{Civic Interconnect, Ely, MN, USA}

\date{\today}

\begin{document}

\maketitle
\vspace{-1em}

% Explicitly mark preprint status
\begin{tcolorbox}[colback=gray!10, colframe=black!20, boxrule=0.3pt]
    \textbf{Preprint Notice.}
    This preprint has not undergone peer review.
    It is shared to support community discussion, transparency research, and early technical evaluation.
\end{tcolorbox}

\begin{abstract}
    We present a formal ontology of Civic Accountable Entities (CAE),
    defining the kinds of entities that participate in civic accountability relationships.
    CAE partitions entities into six disjoint kinds:
    Actors (A), Sites/Assets (S), Instruments (I), Events (E), Jurisdictions (J), and Observations (O),
    according to their role in obligations, authority relationships, and accountability-bearing exchanges,
    rather than by domain, sector, or function.

    The ontology is explicitly obligation-driven:
    entities are included only insofar as they participate in
    accountability relationships, and
    apparent role changes are modeled through relationships
    rather than reclassification.
    Roles, classifications, and sectoral labels are treated as attributes or relations
    rather than entity kinds, ensuring disjointness and preventing ontological overlap.
    Laws and regulations are modeled as enduring Instruments that ground
    concrete Events without requiring exhaustive modeling of legal texts.

    The six kinds are designed to be stable over time
    and sufficient to represent long-term public investments,
    regulatory regimes, financial flows, and outcome measurements across
    domains including procurement, health, environment, education, and infrastructure.
    CAE provides the ontological foundation for the Civic Exchange Protocol (CEP),
    which models how entities exchange value and authority,
    and for Contextual Evidence and Explanations (CEE),
    which attaches structured explanations to civic decisions.
    By separating what exists from how it moves and how decisions are explained,
    CAE enables interoperable, auditable, and longitudinal analysis
    of public systems while remaining neutral
    with respect to policy positions or causal claims.
\end{abstract}

\begin{keywords}
    Formal ontology,
    civic accountability,
    knowledge representation,
    applied ontology,
    public data interoperability,
    accountability relationships.
\end{keywords}

\input{01_intro}           % Section 1: Introduction and Motivation
\input{02_related}         % Section 2: Related Work
\input{03_principles}      % Section 3: Design Principles and Scope
\input{04_cae}             % Section 4: Ontological Partition 
\input{05_relationships}     % Section 5: Relationships and Structural Constraints
\input{06_eval}          % Section 6: Evaluation via Competency Questions

\input{07_chains}          % Section 7: Laws, Regulations, and Accountability Chains
\input{08_value}           % Section 8: Outcomes, Observations, and Public Value
\input{90_conclusion}      % Section 9, 10: Discussions and Limitations, Conclusion

\appendix
\clearpage

\input{99_appendixA}   % Worked examples
\input{99_appendixB}   % Diagrammatic intuition
\input{99_appendixC}   % Glossary of CAE terms (not category theory)

\bibliographystyle{plainnat}
\bibliography{bib_shared}
\end{document}