% !TeX root = 00P1_cae_ontology.tex

\section{Relationship to CEP and CEE}
\label{sec:relationship}

The Civic Accountable Entities (CAE) ontology is intended to serve as the
foundational object universe for subsequent categorical semantics of the Civic
Exchange Protocol (CEP) and Contextual Evidence and Explanations (CEE).
CAE defines what kinds of entities exist and how they may be related, while CEP and
CEE define how exchanges and evidence are interpreted over those entities.

CEP operates over CAE entities by assigning formal semantics to patterns of
relationships involving Actors, Instruments, Events, and Sites.
Exchanges, obligations, transfers, and authorizations are expressed as structured
compositions of CAE relationships rather than as new ontological constructs.
Because entity kinds and relationship constraints are fixed at the CAE level,
CEP semantics remain uniform across domains and jurisdictions.

CEE builds upon both CAE and CEP by attaching evidence, provenance, and
explanatory structure to Events and Observations grounded in CAE entities.
Explanatory claims and causal hypotheses are expressed relative to the
accountability chains defined by CAE and the exchange semantics defined by CEP.
This separation ensures that evidence and explanation do not require modification
of the underlying ontology.

Together, CAE, CEP, and CEE form a layered framework in which ontological
commitment, exchange semantics, and evidentiary reasoning are clearly separated
yet composable.
