\PassOptionsToPackage{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}{hyperref}
% REQ-FILE: The above line, \PassOptionsToPackage{} must be very first line.

\documentclass[11pt]{article}

% Encoding and font setup for cross-platform reproducibility
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Improve readability for dense theoretical material
\usepackage{setspace}
\onehalfspacing

% Mathematical notation and table formatting for formal semantics
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{bookmark}
\usepackage{graphicx}

% Theorem environments for definitions, lemmas, and formal statements
\usepackage{amsthm}

% plain style is italic body text (for theorems, lemmas, etc.)
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

% definition style = upright body text (for definitions, examples)
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

% remark style = upright, lighter weight (for remarks, notes)
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{note}[theorem]{Note}

% Bibliography management and hyperlink support
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}

% Support multiple author affiliations
\usepackage{authblk}

% Improve typographic quality and line breaking
\usepackage{microtype}

% Categorical and commutative diagrams
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{arrows.meta,calc,fit,positioning,shapes.geometric}

% Callout boxes for examples and emphasis
\usepackage{mdframed}
\usepackage[most]{tcolorbox}

% Keywords macro for structured abstract metadata
\providecommand{\keywords}[1]{\textbf{Keywords: } #1}

% Notation for Raw vs Canonical categories in CEP/CEE
\newcommand{\Raw}{\mathsf{Raw}}
\newcommand{\Canon}{\mathsf{Canon}}

% Reusable figure callout box for visual emphasis
\newcommand{\FigureCallout}[2]{%
  \begin{tcolorbox}[
    colback=gray!5,
    colframe=black!40,
    title={#1},
    fonttitle=\bfseries,
    arc=3pt,
    boxrule=0.5pt,
    width=\linewidth,
    enhanced,
    breakable
  ]
  #2
  \end{tcolorbox}
}

% Paragraph spacing prioritizes readability over traditional indentation
\setlength{\parskip}{0.75em}
\setlength{\parindent}{0em}

% Compact list formatting for dense conceptual content
\usepackage{enumitem}
\setlist[itemize]{itemsep=0.2em, topsep=0.2em, parsep=0em, partopsep=0em}

% Defensive re-definition ensures commands exist if loaded earlier
\providecommand{\Raw}{\mathrm{Raw}}
\providecommand{\Canon}{\mathrm{Canon}}

\title{A Bicategorical Semantics of Civic Explanations and Vertical Domains}

\author[1,2]{Denise Case}
\affil[1]{Northwest Missouri State University, Computer Science and Information Systems, Maryville, MO, USA}
\affil[2]{Civic Interconnect, Ely, MN, USA}

\date{\today}

\begin{document}

\maketitle
\vspace{-1em}

% Explicitly mark preprint status
\begin{tcolorbox}[colback=gray!10, colframe=black!20, boxrule=0.3pt]
  \textbf{Preprint Notice.}
  This preprint has not undergone peer review.
  It is shared to support community discussion, transparency research,
  and early technical evaluation.
\end{tcolorbox}

\begin{abstract}
  The Civic Exchange Protocol (CEP) provides a canonical substrate for
  civic entities, relationships, and exchanges.
  The Contextual Evidence and Explanations (CEE) layer extends this
  substrate with structured, evidence-based explanations of why
  particular civic decisions, alerts, or highlights occur.
  This paper develops a geometric and categorical perspective on the
  combined CEP+CEE architecture.
  We introduce the notion of a \emph{vertical domain} as a structured
  slice through the global civic graph, and show how verticals compose
  into a stack with a bicategorical semantics:
  objects as entities, 1-morphisms as relationships plus provenance,
  and 2-morphisms as explanations that relate one morphism-level
  interpretation to another.
  The resulting framework is mathematically tidy yet operationally
  useful: it supports interoperable modeling across domains such as
  SME-friendly procurement, community asset access, environmental
  compliance, education access, and disaster resilience, while making
  the logic of explanations explicit and reusable.
\end{abstract}

\begin{keywords}
  Civic data;
  category theory;
  functorial data modeling;
  interoperability;
  canonicalization;
  identifiers;
  provenance.
\end{keywords}

\input{01_intro}       % Introduction
\input{02_background}  % Background and preliminaries
\input{03_cep}         % Recap of CEP canonical layer
\input{04_verticals}   % Vertical domains
\input{05_stack}       % Stack of vertical domains
\input{06_bicat}       % Bicategorical semantics
\input{06_maps}        % Maps between vertical domains
\input{07_cases}       % Illustrative case studies
\input{08_related}     % Related work
\input{09_agenda}      % Research agenda
\input{90_conclusion}  % Limitations, future work, conclusion

\appendix
\clearpage

\input{99_appendixA}   % Category background
\input{99_appendixB}   % Worked examples
\input{99_appendixC}   % Proof sketches
\input{99_appendixD}   % Diagrammatic intuition
\input{99_appendixE}   % Categorical Concepts for CEE
\input{99_appendixF}   % Supporting categorical terminology

\bibliographystyle{plainnat}
\bibliography{bib_shared}
\end{document}
